---
title: "EDA Project - Randy Crane"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning=FALSE)
```

## Introduction

I have had this data set for about 2 years, and I'm looking forward to finally being able to explore it in this project now that I have these tools. The data set was created by TouringPlans.com specifically to give programmers and statisticians who may be interested in working with them someday some actual data they have compiled to practice on.

A total of four data sets were provided, each consisting of wait time data for one of the four Walt Disney World theme parks (Magic Kingdom, Epcot, Hollywood Studios, and Animal Kingdom) for a particular time period. I have chosen to use the Magic Kingdom ("MK") data set for this project.

To keep the project scope from getting away from me, before importing the set into R, I did some cleaning on it--specifically, I narrowed the date range down to the most recent two complete years for which data was provided (2015 and 2016) and removed some columns that were included but had no data.

For each data, the average weight time in minutes ("SPOSTMIN") of all the attractions in the park was calculated several times a day. Each of these observations is recorded. The purpose of this project is to perform an exploratory data analysis on these wait times to look for possible patterns and to see what factors (e.g. number of hours the park is open, number of hours other parks are open, mean minimum temperature for the day, season, etc.) have a noticeable impact on those wait times. 

There are some quirks and limitations to the data set that I wish did not exist. The two primary ones are that, for some reason, Epcot's park hours were not included, and instead of the mean temperature for each day, the mean <i>minimum</i> temperature for each day was provided. I think the mean temperature would be more useful, but the mean minimum is better than nothing.  


<p>&nbsp;</p>
<p>&nbsp;</p>
Now, let's load the data set & packages and get a look at the structure.

```{r Loading data set}
getwd()
setwd("D:\\Documents\\School - WGU\\Term 4\\C751\\Project")

mk <- read.csv("MK04.csv")

library(dplyr)
library(ggplot2)
library(GGally)
library(memisc)
library(gridExtra)
library(RColorBrewer)
library(scales)
library(tidyr)

str(mk)
```

<p>&nbsp;</p>
## Analysis and Exploration

This data set consists of 73,222 observations of 24 variables, for a total of 1,757,328 data points.

Not all of these variables are relevant to a summary, so here is the summary with only those that are. These may or may not all end up being useful later in the analysis. 

```{r summary_of_relevant_numeric_columns}
cols <- c("SPOSTMIN", "WGT", "MKHOURS", "HSHOURS", "AKHOURS", "IAHOURS", "UFHOURS", "WDWMINTEMP_MEAN");
summary(mk[cols])
```

It's time to start looking at some visualizations of the data. We'll begin with a simple histogram showing the distribution of wait times.
```{r Visualization1_Distribution_of_wait_times_histogram}
ggplot(aes(x = SPOSTMIN), data = mk) +
  geom_histogram(binwidth = 5)
```

This is somewhat surprising. Considering what I thought I knew of theme park wait times, and considering the summary shown above, I would have expected a more normal distribution centered around 36 minutes. Instead, we have a positively skewed distribution with the vast majority being in the 5-minute range. There are definitely some outliers, too, and I'm sure we'll see those again. Given what I have personally experienced in theme park wait times, those outliers are legitimate data,so generally I will keep them included.

Now, we'll get a snapshot of many of the data pairs.

```{r Visualization2_Matrix_of_Plots}
cols <- c("SPOSTMIN","WDW_TICKET_SEASON" , "DAYOFWEEK", "WEEKOFYEAR", "SEASON", "HOLIDAYPX", "HOLIDAYM", "INSESSION", "MKHOURS", "HSHOURS", "AKHOURS", "IAHOURS", "UFHOURS", "WDWMINTEMP_MEAN", "CAPACITYLOST_MK");

ggpairs(mk[cols], 
        lower = list(continuous = wrap("points", shape = I('.'))), 
        upper = list(combo = wrap("box", outlier.shape = I('.'))),
        cardinality_threshold = 20)
```

<i>This is a lot of visuals in a relatively small space. <a href="https://github.com/belgarion42/c751/blob/master/GGPairs-Plot.png" target="_blank">A larger version is available here.</a></i> 

This is a good quick snapshot of many of the possible data pairs. Looking at the correlation coefficients, some non-correlations are obvious and expected (Magic Kingdom park hours and the mean minimum temperature for a particular day are completely independent of one another, and so the correlation coefficient of -0.009 is expected.) 

However, there are also some stronger correlations. Since wait times are the primary variable of interest, as I look at this visualization, I see a moderately strong correlation between wait times and these other variables:

<ul>
<li>Holiday Metric (HOLIDAYM) ... 0.311</li>
<li>Percentage of Schools in Session (INSESSION) ... -0.305</li>
<li>Total Open Hours for Magic Kingdom (MKOPEN) ... 0.302</li>
</ul>

Surprisingly, the two strongest correlations are between wait times and Universal Studios' Islands of Adventure total open hours (0.341) and between wait times and Universal Studios total open hours (0.320).

Let's take a closer look at some individual plots to see what we can learn, starting with a box plot of wait times by WDW_TICKET_SEASON. 

```{r Visualization3_Box_Plot_of_wait_times_by_WDW_TICKET_SEASON}
ggplot(aes(x = WDW_TICKET_SEASON, y = SPOSTMIN), data = mk) +
  geom_boxplot()
```

There are a couple of things I notice right away about these box plots. First, as expected, the mean wait times are longest during peak season and decrease from there. Somewhat surprisingly, peak season has a lot of outliers, but considering that "peak season" dates include Spring Breaks, Summer, Christmas, and other holidays, I expect a lot of variability in that one. The outliers in regular season may warrant more scrutiny. We may be able to see this with a scatterplot of wait times vs. month of the year.

```{r Visualization4_Scatterplot_of_wait_times_vs_month_of_the_year_with_jitter.}
ggplot(aes(x=(MONTHOFYEAR),y=SPOSTMIN, color = WDW_TICKET_SEASON), data=mk) +
  geom_jitter(alpha = 1/4) +
  scale_x_continuous(breaks = seq(1, 12, 1))
```

Sure enough, the season varies from month to month for much of the year. Only June and October are each consistently one season (peak and regular, respectively). Also, there are relatively few outliers--though the ones at the bottom of the plot are confusing and need more scrutiny, just perhaps not in this project.

```{r Visualization5_}
ggplot(aes(x = DAYOFWEEK, y = SPOSTMIN, color = WDW_TICKET_SEASON), data=mk) +
  geom_jitter(alpha = 1/4) +
  scale_x_continuous(breaks = seq(1, 7, 1), labels=c("1" = "Sunday", "2" = "Monday", "3" = "Tuesday", "4" = "Wednesday", "5" = "Thursday", "6" = "Friday", "7" = "Saturday"))
```

This time, we are looking at the wait times by day of the week, again color-coded by ticket season.These findings definitely surprise me. I would expect both weekend days days to be the busiest, but Sunday is noticeably lower than the other days. Let's take a look at the summary for each day to find out more.

```{r summary_for_days_of_week}
by(mk$SPOSTMIN, mk$DAYOFWEEK, summary)
```

We can see from this table that Sunday is, in fact,  the lowest (or tied for the lowest) at every quantile level, and the mean is the lowest. It can be hard to really see how all 7 days compare on a table like this, though. Let's visualize it as a box plot.

```{r Visualization6_}
ggplot(aes(x = DAYOFWEEK, y = SPOSTMIN, group = DAYOFWEEK), data = mk) +
  geom_boxplot() +
  stat_summary(fun = mean, geom = "point", shape = 4)
```

Sure enough, Sunday's median and mean are lower, as are the quartiles. If we remove the outliers, this becomes even more obvious.

```{r Visualization7_}
ggplot(aes(x = DAYOFWEEK, y = SPOSTMIN, group = DAYOFWEEK), data = mk) +
  geom_boxplot() +
  stat_summary(fun = mean, geom = "point", shape = 4) +
  ylim(0, quantile(mk$SPOSTMIN, 0.95))
```

Sunday's median  is lower than 3 days, and the same as the other 3. The quartiles for Sunday are definitely smaller. And the mean is lower than all the other days. Why would this be? Wouldn't weekend days be highest?

Not necessarily. WDW is a destination vacation, not generally a locals' park (like Disneyland is). Because it is a destination, people are more likely to arrive on the weekend. In my experience, guests often like to finish up their packing and preparations on Saturday, then travel Sunday and stay 7 nights. These stays are Sunday - Sunday.

That makes Sunday the arrival/departure day for a larger-than-average percentage of guests, which then also makes it the day they are least likely to visit the theme parks. If we had observations for arrival/departure day, we could confirm this, but that is my educated guess.

I would like to explore this further. Let's see if the daily attendance pattern holds across ticket seasons.

```{r Visualization8_box_plots_for_DAYOFWEEK}
p1 <- ggplot(aes(x = DAYOFWEEK, y = SPOSTMIN, group = DAYOFWEEK), data = subset(mk, WDW_TICKET_SEASON == "peak")) +
  geom_boxplot(color = "red") +
  ylim(0, quantile(mk$SPOSTMIN, 0.95)) +
  scale_x_continuous(breaks = seq(1, 7, 1), labels=c("1" = "Sunday", "2" = "Monday", "3" = "Tuesday", "4" = "Wednesday", "5" = "Thursday", "6" = "Friday", "7" = "Saturday")) +
  xlab("DAYOFWEEK - Peak Season") +
  stat_summary(fun = mean, geom = "point", shape = 4)

p2 <- ggplot(aes(x = DAYOFWEEK, y = SPOSTMIN, group = DAYOFWEEK), data = subset(mk, WDW_TICKET_SEASON == "regular")) +
  geom_boxplot(color = "green") +
  ylim(0, quantile(mk$SPOSTMIN, 0.95)) +
 scale_x_continuous(breaks = seq(1, 7, 1), labels=c("1" = "Sunday", "2" = "Monday", "3" = "Tuesday", "4" = "Wednesday", "5" = "Thursday", "6" = "Friday", "7" = "Saturday")) +
  xlab("DAYOFWEEK - Regular Season") +
  stat_summary(fun = mean, geom = "point", shape = 4)

p3 <- ggplot(aes(x = DAYOFWEEK, y = SPOSTMIN, group = DAYOFWEEK), data = subset(mk, WDW_TICKET_SEASON == "value")) +
  geom_boxplot(color = "blue") +
  ylim(0, quantile(mk$SPOSTMIN, 0.95)) +
  scale_x_continuous(breaks = seq(1, 7, 1), labels=c("1" = "Sunday", "2" = "Monday", "3" = "Tuesday", "4" = "Wednesday", "5" = "Thursday", "6" = "Friday", "7" = "Saturday")) +
  xlab("DAYOFWEEK - Value Season") +
  stat_summary(fun = mean, geom = "point", shape = 4)

grid.arrange(p1, p2, p3, ncol = 1)
```

There is definitely more variation between seasons. 

```{r_summary_of_DAYOFWEEK_by_WDW_TICKET_SEASON}
season_peak <- subset(mk, WDW_TICKET_SEASON == "peak")
season_regular <- subset(mk, WDW_TICKET_SEASON == "regular")
season_value <- subset(mk, WDW_TICKET_SEASON == "value")

by(season_peak$SPOSTMIN, season_peak$DAYOFWEEK, summary)
by(season_regular$SPOSTMIN, season_regular$DAYOFWEEK, summary)
by(season_value$SPOSTMIN, season_value$DAYOFWEEK, summary)
```

Having 3 summaries of 7 days each is a little hard to follow. Still, as we view these three summaries, we can see that, while the mean remains lowest on Sundays across the seasons, the median is lowest for Sunday only in peak season. I'm not sure why that is. We may come back and try to understand it later. 

For now, there are other aspects of this data set I want to explore, so let's move on.

```{r Visualization9_Scatterplot_to_compare_wait_times_to_mean_min_temps}
ggplot(aes(x = WDWMINTEMP_MEAN, y = SPOSTMIN), data = mk) +
  geom_jitter(alpha = 1/20) +
  ylim(0, quantile(mk$SPOSTMIN, 0.95))
```

Here we are viewing the wait times vs. the mean minimum temperatures for each day. 

Well, that doesn't look right. Assuming WDWMINTEMP_MEAN is what it sounds like it is, this means that the days with the highest mean minimum temperature have the longest wait times. That wouldn't be too surprising, except this is central Florida, and a mean <i>minimum</i> temperature of 70° - 75° indicates <a href="https://www.weather-us.com/en/florida-usa/orlando-climate#temperature" target="_blank">a mean high temperature of 90°+</a>. Why would these days have the longest wait times? Wouldn't they be the days people would want to avoid?

Let's validate this data by comparing the WDWMINTEMP_MEAN to the month of the year to see which months are the highest minimum temperatures. 

```{r Visualization10_comparing_temperatures_to_months}
ggplot(aes(x = MONTHOFYEAR, y = WDWMINTEMP_MEAN, color = WDW_TICKET_SEASON), data = mk) +
  geom_jitter() +
  scale_x_continuous(breaks = seq(1, 12, 1)) +
  ylim(45, 80)
```

Ah, OK, this does make some sense now. Those months with the highest temperatures also have the highest concentration of value season days. A lot of guests are willing to brave the hot weather to pay the lower price. So it seems the real driving factor here is not the temperatures, but what months / ticket seasons they fall in.

That was interesting, but there is much more to see and explore.
```{r Visualization11_HOLIDAYM}
ggplot(aes(x = HOLIDAYM, y = SPOSTMIN, group = HOLIDAYM), data = mk) +
  geom_jitter(alpha = 0.25, color = "lightsteelblue2") +
  geom_boxplot(alpha = 0.25) +
  stat_summary(fun = mean, geom = "point", shape = 4) +
  scale_x_continuous(breaks = seq(0, 5, 1))
```

This is unexpected. Here we have a visualization of wait times vs. the "HOLIDAYM" metric, which is a number that indicates a given dates proximity to a holiday combined with how "major" that holiday is considered when it comes to travelers. The fact that there are relatively few data points in the 5 (highest) HOLIDAYM surprises me, but that may be due to the difficulty in gathering the data at major holiday times.

What is surprising to me is the lack of any predictable pattern. I would have expected the wait times to either increase more-or-less consistently from 0 - 5 or to decrease in that same way, but it does neither. 0 is the lowest and 5 is the highest, but in between comes 2, 1, 4, and 3, in that order. I really have no idea why this is.

Considering the fact that the data pairs matrix showed a correlation coefficient of 0.311, I'm even more surprised to see apparent unpredictablity, as HOLIDAYM should be a moderately strong predictor of wait times. Unfortunately, I don't think I have enough data to explain this in more detail. We may return to it later, though.

```{r Visualization12_INSESSION}
ggplot(aes(x = INSESSION, y = SPOSTMIN), data = mk) +
  geom_point(alpha = 0.1) +
  geom_smooth(color = "blue")
```

Interesting and again not as expected. One would assume that the more schools are not in session (an INSESSION number closer to 0), the higher the average wait time. But that doesn't appear to be the case here. HOwever, there is so much data, there is a great deal of overplotting, so let's see what we can do with that.

```{r Visualization13_INSESSION_reducing_overplotting}
ggplot(aes(x = INSESSION, y = SPOSTMIN), data = mk) +
  geom_jitter(alpha = 0.005) +
  coord_cartesian(ylim = c(0, quantile(mk$SPOSTMIN, 0.95))) +
  stat_summary(fun ="mean",  geom="pointrange", color = "red") +
  stat_smooth(geom="line", color = "blue", size = 2) +
  facet_wrap(~WDW_TICKET_SEASON) +
  theme_bw()
```

That's better. I had really expected a stronger relationship here, given the relatively high (actually moderate, but higher than most others) inverse correlation. 

From what I see here, it seems that during peak season, the mean wait times are within a fairly narrow range (40-50 minutes) until we get to about 65% of schools in session. This seems to show that during the peak season, while most schools are not in session, the percentage of schools in session doesn't have much of an impact. Then we see a bump--which may be related to </i>which</i> schools are still not in session and may indicate a desire of those families to get in a trip when most kids are back in school. Then, predictably, it drops off dramatically.

During the regular season, the impact of schools being in session is more clear, and more what I would expect. For value season, there are so few data points of schools not being in session that the graph is inconclusive. 

```{r Visualization14_INSESSION_reducing_overplotting_with_and_without_seasons}
p1 <- ggplot(aes(x = INSESSION, y = SPOSTMIN), data = mk) +
  geom_jitter(alpha = 0.005) +
  coord_cartesian(ylim = c(0, quantile(mk$SPOSTMIN, 0.95))) +
  stat_summary(fun ="mean",  geom="pointrange", color = "red") +
  stat_smooth(geom="line", color = "blue", size = 2) +
  facet_wrap(~WDW_TICKET_SEASON) +
  theme_bw()

p2 <- ggplot(aes(x = INSESSION, y = SPOSTMIN), data = mk) +
  geom_jitter(alpha = 0.005) +
  coord_cartesian(ylim = c(0, quantile(mk$SPOSTMIN, 0.95))) +
  stat_summary(fun ="mean",  geom="pointrange", color = "red") +
  stat_smooth(geom="line", color = "blue", size = 2) +
  theme_bw()

grid.arrange(p1, p2, ncol = 1)
```

If we stack these on top of each other, we can see that the combined chart very closely mirrors the peak season chart. 

So far we've been looking at the park as a whole. What if we explore wait times subset by NEWNESS? Do attractions that have had major refurbishments recently have longer median or mean wait times?
```{r Visualization15_by_newness}
#
newness_groups <- group_by(mk, NEWNESS)
mk.newness_by_groups <- summarise(newness_groups,
          wait_time_mean = mean(SPOSTMIN),
          wait_time_median = median(SPOSTMIN),
          n = n())

ggplot(aes(x= NEWNESS, y = wait_time_mean), data = mk.newness_by_groups) +
  geom_point(fill = "yellow", size = 4, shape = 23) +
  geom_point(aes(x= NEWNESS, y = wait_time_median, color = "green", size = 4), data = mk.newness_by_groups) + 
  guides(color = FALSE, size = FALSE) +
  ylab("Wait Times (Mean and Median)") +
  xlab("Refurbishments") +
  scale_x_discrete(breaks = c("MAJOR REFURB WITHIN 18 MONTHS", 
                              "MINOR REFURB WITHIN 18 MONTHS", 
                              "MINOR REFURB WITHIN 30 DAYS", 
                              "MINOR REFURB WITHIN 6 MONTHS", 
                              "MINOR REFURB WITHIN 7 DAYS"), 
                   labels = c("Major w/in 18 months", 
                              "Minor w/in 18 months", 
                              "Minor w/in 30 days", 
                              "Minor w/in 6 months", 
                              "Minor w/in 7 days"))
  

ggplot(aes(x= NEWNESS, y = SPOSTMIN), data = mk) +
  geom_point(alpha = 1/5, position = position_jitter(h = 0)) +
  geom_line(stat = "summary", fun = "mean", color = "red") +
  geom_line(stat = "summary", fun = quantile, fun.args = list(probs = 0.5), color = "blue")

```

This plot shows the same general trend for both the mean (yellow) and median (red)--the longest  wait times are for attractions having a minor refurbishment within 6 months. The only reason I can think of for this is that the attractions with minor refurbishments within 6 months are also the "headliners" and/or guest favorites and so they naturally generate higher wait times. Without a list of what attractions are in each category, though, I can't say this for certain.  

A major refurbishment within 18 months appears to have no effect.

That was interesting, but not too revealing. Let's see what happens when we look at wait times vs. opening hours for each park (except Epcot, which I have no data for), including the two Universal Studios parks.

```{r Visualization16_Compare_to_total_opening_hours_for_each_park}
#Compare to total opening hours for each park (multiple charts)


```

Explanation & observations of Viz 16.
```{r Visualization17_}
#Compare to season (not ticket season)

```

Explanation & observations of Viz 17.
```{r Visualization18_}
#Can timepart be used to show wait times by time of day (maybe make a factored column using it)?


```

Explanation & observations of Viz 18.
```{r Visualization19_}


```

Explanation & observations of Viz 19.
```{r Visualization20_}


```

Explanation & observations of Viz 20.

<p>&nbsp;</p>
## Final Plots and Summary

```{r Final Plots}


```


```{r Building_the_Linear_Model}
m1 <- lm(I(SPOSTMIN) ~ I(WDW_TICKET_SEASON), data = mk)
m2 <- update(m1, ~ . + WDWMINTEMP_MEAN)
m3 <- update(m2, ~ . + HOLIDAYM)
m4 <- update(m3, ~ . + INSESSION)
m5 <- update(m3, ~ . + MKHOURS)
mtable(m1, m2, m3, m4, m5)
```

<p>&nbsp;</p>
## Reflection



